---
title: "Seattle Pet Names"
author: "Alejandra Gerosa"
date: "August 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(devtools)
library(choroplethrZip)
library(ggmap)
register_google(key = Sys.getenv("google_maps_key"))
```

## Tidy Tuesday - Seattle Pet names

Taking a first stab at Tidy Tuesday!

```{r load}
seattle_pets <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-26/seattle_pets.csv")

```

## Explore
Take a look:

Let's make top 10 tables:

```{r explore}
summary(seattle_pets)
table(seattle_pets$zip_code)

top_ten_names <- seattle_pets %>%
  group_by(animals_name) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10)

top_ten_primary_breed <- seattle_pets %>%
  group_by(primary_breed) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10)

top_ten_secondary_breed <- seattle_pets %>%
  group_by(secondary_breed) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10)

top_ten_zip_code <- seattle_pets %>%
  group_by(zip_code) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10)

```

Let's do a chart

```{r plots}

ggplot(top_ten_names, aes(animals_name, count)) +
  geom_col()


```

## Try a map

First I try a choropleth because that's where google landed me. 

```{r map}
for_map <- seattle_pets %>%
  group_by(region = zip_code) %>%
  summarize(value = n())

for_map %>% 
  filter(region == "98115") 
zip.map %>%
  filter(region == "98115")


#%>% 
#  zip_choropleth(county_zoom = 53055)


```

Something isn't working *but* the choropleths package does have exactly what I needed: a way to match zipcodes to a lat-long pair. So, first I need to add that information to my dataset.

```{r latlong}
sp_geo <- seattle_pets %>% 
  left_join(select(zip.map, long, lat, region), by = c("zip_code" = "region")) %>%
  group_by(license_issue_date, license_number, animals_name, species, primary_breed, secondary_breed, zip_code) %>% 
  summarize(long = mean(long, na.rm = TRUE), lat = mean(lat,na.rm = TRUE))

```
And now we make a map

```{r map}
seattle <- c(lon = mean(sp_geo$long, na.rm = TRUE), lat = mean(sp_geo$lat, na.rm = TRUE))
seattle_map <- get_map(seattle, zoom = 12)
sp_for_map <- sp_geo %>%
  group_by(long, lat) %>% 
  summarise(count = n())

ggmap(seattle_map) +
  geom_point(aes(long, lat, size = count), data = sp_for_map)

```